---
title: "Weather Events and their Consequences"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Synopsis

This analysis uses NOAA Storm Data to answer two basic questions about severe weather events and their consequences.  Specifically, this analysis answers these two questions:

1. Across the United States, which types of events (as indicated in the ```EVTYPE``` variable) are most harmful with respect to population health?
2. Across the United States, which types of events have the greatest economic consequences?

This document has two main sections.  First, a data processing section describes how to obtain and process the data before analysis.  A results section follows that analyzes the data to answer the two questions.

## Data Processing

The first step is to download, extract, and read the data file.

```{r download, cache = TRUE}
summary(cars)
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
